From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study

### 2025-07-27 draft_v1/draft_v2.tex
update draft
draft edit_Vitorfeedback

\date{2025-06-27}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates how the internal structure of
depression--precariousness dynamics shapes responsiveness to external
support. Using cross-sectional data from the HELIUS study, we first
apply cycle-capable causal discovery algorithms to estimate directional
dependencies between financial stress, domain-specific precariousness,
and depressive symptoms. Our multi-resolution analysis---spanning both
aggregate and symptom-level constructs---highlights financial stress as
a plausible upstream driver of depression, with symptoms such as sleep
disturbance and depressed mood emerging as initiators of broader social
vulnerability. To translate these structural insights into dynamic
predictions, we construct a family of analytically calibrated linear
stochastic models. Simulations reveal that variations in internal
structure can substantially alter system responses to financial stress
reduction. Some models show rapid but transient improvements in
depression, while others adjust more slowly yet sustain gains over time.
Precariousness tends to respond more immediately across all
configurations, though its trajectory also shifts subtly with structural
variation. Together, these findings demonstrate that intervention
outcomes depend not only on the magnitude of external input but also on
latent properties of the system. More broadly, this study illustrates
how combining causal discovery with dynamical modeling supports
principled, simulation-based exploration of intervention
scenarios---grounded in data, yet able to probe counterfactual
structure. This approach offers a scalable framework for diagnosing
hidden system constraints and informing targeted public health
strategies in complex psychosocial domains.
\end{abstract}

@@

Mental health disorders represent a growing global health concern,
especially in urbanized regions (Gruebner et al., 2017; World Health
Organization, 2022). In 2019, one in eight people worldwide were living
with a mental health condition, with the highest disability-adjusted
life years (DALYs) due to mental and addictive disorders concentrated in
high-income countries such as those in Northern Europe, North America,
and Australia (Rehm \& Shield, 2019; World Health Organization, 2022).
Despite growing awareness, governments have struggled to design
effective responses. Mental health outcomes arise from complex,
multi-level dynamics---intertwined with the social, economic, and
spatial structures of urban environments---making both diagnosis and
intervention design deeply challenging (Van Der Wal et al., 2021).

Recent work has shifted focus from individual-level vulnerability to
broader social and structural determinants. While high-income nations
show elevated overall DALYs, within-country disparities reveal that
income inequality is a significant predictor of mental health burden
(Rehm \& Shield, 2019). Moreover, precarious working conditions, housing
instability, and neighborhood disadvantage have all been linked to
increased risk of depression and anxiety (Fone et al., 2014; Pevalin et
al., 2017; Rönnblad et al., 2019; Rugulies et al., 2023). This growing
literature highlights that poor mental health is not merely a personal
issue, but one rooted in sustained socioeconomic stress and structural
uncertainty.

Building on this, recent studies have introduced the concept of
precariousness as a multidimensional condition characterized by
instability and lack of control across several life domains---including
employment, finances, housing, social relations, and cultural belonging
(Elsenburg et al., 2025; McKee et al., 2017). This broader view reveals
how different forms of insecurity can co-occur and compound, forming an
ecosystem of risk. However, despite growing evidence of association
between precariousness and mental health, a fundamental question remains
unresolved: \emph{How do these factors causally influence one another?}
Most existing studies rely on cross-sectional correlations, leaving open
questions about directionality and feedback. Without stronger causal
insight, it remains difficult to identify priority targets for
intervention or to anticipate unintended effects.

This study aims to investigate the causal mechanisms linking
precariousness and depression---and how those mechanisms shape system
responses to external intervention. Specifically, we ask: \emph{How does
the internal configuration of the precariousness--depression
relationship affect the outcomes of stress-reducing interventions?} To
answer this, we combine two approaches. First, we use cycle-capable
causal discovery algorithms to infer directional relationships among
domain-specific precariousness factors, and depressive symptoms. By
examining both aggregated measures (depression sum score and a composite
precariousness index) and disaggregated variables (individual depressive
symptoms and specific life-domain indicators), we capture broad patterns
of relationship as well as fine-grained causal pathways. Second, we
translate these insights into a computational dynamical model that
simulates how depression and precariousness co-evolve under varying
levels of external stress. This model enables us to probe how system
properties---such as feedback strength and stochastic noise---govern
responsiveness to intervention. In some configurations, the system
exhibits bistability, remaining stuck in a high-risk state unless
external support exceeds a critical threshold. In others, it shows
unistable behavior, responding more gradually to incremental changes. By
integrating causal discovery with dynamic simulation, we aim to
demonstrate how the internal structure of the depression--precariousness
system shapes its susceptibility to change---and, by extension, how
intervention strategies might be tailored to the architecture of
underlying vulnerability.

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study,
which has been described in detail elsewhere (Snijder et al., 2017;
Stronks et al., 2013). It captures the diverse population of the city of
Amsterdam by including the six main ethnic groups and provides
comprehensive health and lifestyle data, including depressive symptoms
as measured by the PHQ-9 (Galenkamp et al., 2017). To operationalize
indicators of precariousness, we draw on the framework outlined in
previous research (Elsenburg et al., 2025) and select a set of relevant
variables. To ensure a robust representation of precariousness in our
causal discovery models, we conducted exploratory analyses to identify
consistent and meaningful data-driven factor structures. These analyses
led to the identification of five distinct precariousness
factors---three domain-based and two reflecting recent stressors---each
composed of multiple observed variables, as outlined below. Additional
details on the HELIUS study and the factor extraction process are
provided in the \hyperref[sec-appendix]{Appendix}.

@@

After preprocessing, the HELIUS dataset includes 21,628 individuals. In
addition to the five identified precariousness factors, we computed an
overall precariousness score by by summing these five factor scores,
thereby capturing a composite measure of accumulated precariousness
across domains. Depression is represented using the PHQ-9 (Kroenke et
al., 2001), both as a total sum score and as individual symptom scores.
In the subsequent causal discovery analysis, we explore the relationship
between depression and precariousness using both aggregated scores and
their disaggregated components. See Figure~\ref{fig-dist} for the
overall distributions of the variables used in the analysis.

@@

To uncover directional relationships between precariousness and
depression, we use constraint-based causal discovery algorithms that
allow for feedback loops and latent confounding. Specifically, we apply
the \emph{Fast Causal Inference} (FCI) and \emph{Cyclic Causal
Inference} (CCI) algorithms, both of which can detect cycles and
non-acyclic structures (Mooij \& Claassen, 2020; Strobl, 2019). For
reference, we also include the \emph{PC} algorithm, one of the most
widely used methods for acyclic causal discovery (Spirtes et al., 2001).
While our full analysis includes results from three causal discovery
algorithms---FCI, CCI, and PC---we focus on FCI in the main text, as it
offers a strong balance between flexibility and theoretical soundness.
FCI accommodates latent confounding and potential feedback, making it
well-suited for the complexity of our data (Mooij \& Claassen, 2020).
The PC algorithm is included for reference but relies on more
restrictive assumptions, such as acyclicity and no unmeasured
confounding. CCI, while capable of modeling cycles explicitly, comes
with certain theoretical limitations. Results from both PC and CCI are
provided in the \hyperref[sec-appendix]{Appendix}
(Section~\ref{sec-cci}, Section~\ref{sec-pc}) for completeness and
comparison. For readers interested in detailed explanations of the
algorithms, edge interpretations, and graph types (e.g., PAG, MAAG,
CPDAG), as well as algorithmic assumptions and limitations, please refer
to the \hyperref[sec-appendix]{Appendix}
(Section~\ref{sec-causalprimer}).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Within each condition} (e.g., fixed \(\alpha\), threshold, CI
  test, and algorithm), we identify the most frequently occurring edge
  type (e.g., arrowhead, circle, none) across bootstraps for each pair
  of variables (Step 2 in Figure~\ref{fig-workflow}).
\item
  \textbf{Across conditions}, we then determine the most dominant edge
  type based on majority voting. In cases where no edge type is clearly
  dominant (i.e., a tie), we mark the edge as uncertain, represented
  with a dashed line in the final graph (Step 3 in
  Figure~\ref{fig-workflow}).
\end{enumerate}

This summarization process is repeated separately for each causal
discovery algorithm (FCI, CCI, PC), yielding a single robust graph per
algorithm. These final graphs represent only the most stable and
consistent relationships across a wide range of plausible
parameterizations, enhancing both reliability and interpretability. To
support transparency, each final graph is accompanied by a matrix
summarizing the relative frequency of each edge type for every variable
pair---enabling readers to assess the degree of uncertainty or agreement
underlying each connection.

@@

As illustrated in Figure~\ref{fig-concept}, the model includes three
components: depression (\(D\)), represented by the PHQ-9 sum score,
social precariousness (\(P\)), and external financial stress (\(S\)).
Depression and precariousness are modeled as coupled state variables
that influence each other over time, with \(S\) acting as a shared
input. These dynamics are formalized as a pair of linear stochastic
differential equations (SDEs):

\[
\begin{aligned}
dP &= \lambda_P \left(\alpha_{PS} S + \alpha_{PD} D - P \right) dt + \sigma_{P} dW_{P} \\
dD &= \lambda_D \left(\alpha_{DS} S + \alpha_{DP} P - D \right) dt + \sigma_{D} dW_{D} 
\end{aligned}
\]

Here, \(D\) denotes depression, \(P\) denotes social precariousness, and
\(S\) is the external stressor associated with financial stress. The
parameters \(\lambda_D\) and \(\lambda_P\) control the timescales over
which each variable adjusts toward its input-driven target state. The
coefficients \(\alpha_{PS}\) and \(\alpha_{DS}\) quantify the direct
effects of the stressor on \(P\) and \(D\), while \(\alpha_{DP}\) and
\(\alpha_{PD}\) represent the cross-influences between depression and
precariousness. Stochastic fluctuations are introduced through the noise
terms \(\sigma_P, dW_P\) and \(\sigma_D, dW_D\), with \(dW_P\) and
\(dW_D\) representing independent Wiener processes. This linear
structure offers a transparent and analytically tractable framework for
examining how variations in external financial stress influence the
population-level dynamics of depression and social precariousness.
Crucially, the model's simplicity enables direct calibration from
empirical data: all parameters can be derived analytically from observed
variances and covariances in the HELIUS cohort. The following
Section~\ref{sec-calibration} details the procedure used to perform this
calibration.

@@

\includegraphics[width=0.4\textwidth,height=\textheight]{img/conceptmodel.png}

@@

\caption{\label{fig-concept}High-level conceptual model of causal
structure linking social precariousness (P.soc) and depression (PHQsum),
influenced by external financial stress (S.fin).}

@@

We fix the adjustment rates \(\lambda_D = \lambda_P = 1\), which
normalizes the time scale given the cross-sectional nature of the HELIUS
data. To resolve the remaining degrees of freedom, we treat the feedback
from precariousness to depression (\(\alpha_{DP}\)) as a free parameter,
and solve algebraically for all other coefficients --- including the
direct effects of stress (\(\alpha_{DS}, \alpha_{PS}\)), the reverse
feedback from depression to precariousness (\(\alpha_{PD}\)), and the
noise terms (\(\sigma_D, \sigma_P\)). These solutions are expressed as
closed-form functions of the empirical covariances: \(\mathrm{Var}(D)\),
\(\mathrm{Var}(P)\), \(\mathrm{Var}(S)\), \(\mathrm{Cov}(D, P)\),
\(\mathrm{Cov}(D, S)\), and \(\mathrm{Cov}(P, S)\).

This structure creates a family of analytically calibrated models, each
consistent with the same observed covariances but differing in their
internal configuration. However, we constrain the space of valid
solutions by imposing two empirical and theoretical conditions: (1) all
structural parameters must be positive, reflecting the assumption that
stress and precariousness contribute adversely to depression; and (2)
the system must be dynamically stable, ensuring convergence to a
stationary distribution. These constraints restrict the allowable range
of \(\alpha_{DP}\) to approximately \([0, 0.698]\) (see Appendix
Section~\ref{sec-analcalibration}).

Figure~\ref{fig-tradeoff} illustrates how the remaining parameters
adjust as \(\alpha_{DP}\) increases to maintain consistency with the
empirical covariances. As \(\alpha_{DP}\) increases, \(\alpha_{PD}\)
must decline to preserve stability, since the product
\(\alpha_{DP} \cdot \alpha_{PD}\) determines the strength of the
feedback loop and must remain below 1. This creates a compensatory
trade-off: as influence from \(P\) to \(D\) strengthens, the influence
from \(D\) to \(P\) weakens. Meanwhile, the external coupling terms ---
\(\alpha_{DS}\) and \(\alpha_{PS}\) --- also shift to absorb variance in
a way that maintains alignment with the empirical data. Notably, the
noise amplitudes \(\sigma_D\) and \(\sigma_P\) decline slightly as
\(\alpha_{DP}\) increases. This indicates that more of the observed
variability is captured by deterministic interactions, with less left to
be explained by stochastic fluctuations.

@@

In each simulation, financial stress (\(S\)) is proportionally reduced
toward the lowest observed value, modeling a population-wide
intervention applied at varying intensities (0--100\%). For each model,
we simulate 1000 virtual individuals whose baseline stress levels are
sampled from the empirical distribution and scaled according to the
intervention level.

@@

The causal graph in panel (a) highlights key pathways in the
relationships between precariousness factors and depression
(\emph{PHQsum}). Employment precariousness (\emph{P.emp}) and social
precariousness (\emph{P.soc}) are not identified as causes of
depression, whereas financial stress (\emph{S.fin}) appears to play a
causal role. While \emph{P.emp} and \emph{P.soc} are generally not
recognized as causes of other precariousness factors, \emph{S.fin}
emerges as a potential cause, as indicated by its circle edge endpoint.
Supporting this, the proportion matrix plot in panel (b) shows that
\emph{S.fin} has some probability of causally influencing either
\emph{P.emp} or \emph{P.soc.}

@@

Relational stress (\emph{S.rel}) plays a more nuanced role, interacting
with depression and \emph{S.fin} through a latent confounder. Its
connection to \emph{P.soc} is less definitive---the proportion matrix in
Figure~\ref{fig-sum-2} indicates a considerable probability that this
relationship is either absent, ambiguous, or potentially bidirectional.
Meanwhile, housing precariousness (\emph{P.hou}) is not directly
causally related to depression but is linked to \emph{P.emp} and
\emph{S.fin}. While \emph{P.emp} and \emph{S.fin} are identified as
non-causes of \emph{P.hou}, it remains unclear whether \emph{P.hou}
causally influences \emph{P.emp} or if their relationship is mediated by
an unobserved confounder.

@@

@@

\caption{\label{fig-sum}Resulting graph of precariousness factors and
depression sum score using FCI and proportion of edge endpoint types.}

@@

Moving from the sum score representation to the disaggregated
symptom-level graph provides a more granular perspective on the causal
relationships between precariousness factors and depression symptoms.
Unlike the sum score graph, which aggregates all symptoms into a single
measure---potentially obscuring nuanced relationships---the
symptom-level graph reveals the heterogeneity in how specific symptoms
(\emph{con} (concentration), \emph{slp} (sleep), \emph{ene} (energy),
\emph{app} (appetite), \emph{mot} (motor), \emph{sui} (suicidal),
\emph{anh} (anhedonia), \emph{glt} (guilt), and \emph{dep} (depressed
mood)) are influenced by, and in turn influence, different forms of
precariousness.

@@

This discrepancy likely arises from a combination of factors. As the
analysis moves from aggregated to disaggregated variables, the number of
CI tests increases substantially, leading to a loss of statistical
power. With more variables being conditioned on, the effective sample
size per test diminishes, making it harder to detect subtle
dependencies. Over-conditioning---controlling for too many
variables---can also obscure real relationships, resulting in false
negatives. Beyond these statistical power issues, the underlying
assumptions of the two CI tests contribute to their divergence. RCoT is
a nonparametric test that captures both linear and nonlinear
dependencies without assuming a specific distributional form. However,
it is more conservative and requires larger sample sizes, especially in
high-dimensional settings. In contrast, the Gaussian CI test is based on
partial correlations and assumes linear Gaussian relationships, making
it more permissive and more likely to detect weak or spurious
associations. As a result, it is common to see edges where Gaussian CI
suggests a directional relationship, while RCoT indicates no edge. In
the proportion matrix, this appears as a 50--50 tie between absence and
arrowhead symbols. This systematic divergence illustrates the broader
challenge of causal discovery in high-dimensional data: different
statistical assumptions can lead to different conclusions about
whether---and in which direction---variables are causally related.

>>>>

\date{2025-07-26}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates how the internal structure of
depression--precariousness dynamics shapes responsiveness to external
support. Using cross-sectional data from the HELIUS study, we first
apply cycle-capable causal discovery algorithms to estimate directional
dependencies between financial stress, domain-specific precariousness,
and depressive symptoms. Our multi-resolution analysis---spanning both
aggregate and symptom-level constructs---highlights financial stress as
a plausible upstream driver of depression, with symptoms such as sleep
disturbance and depressed mood emerging as initiators of broader social
vulnerability. To translate these structural insights into dynamic
predictions, we construct a family of analytically calibrated linear
stochastic models. Simulations reveal that variations in internal
structure can substantially alter system responses to financial stress
reduction. Some models show rapid but transient improvements in
depression, while others adjust more slowly yet sustain gains over time.
Precariousness tends to respond more immediately across all
configurations, though its trajectory also shifts subtly with structural
variation. Together, these findings demonstrate that intervention
outcomes depend not only on the magnitude of external input but also on
latent properties of the system. More broadly, this study illustrates
how combining causal discovery with dynamical modeling supports
principled, simulation-based exploration of intervention
scenarios---grounded in data, yet able to probe counterfactual
structure. This approach offers a scalable framework for diagnosing
hidden system constraints and informing targeted public health
strategies in complex psychosocial domains.
\end{abstract}

@@

Mental health disorders represent a growing global health concern,
especially in urbanized regions (Gruebner et al., 2017; World Health
Organization, 2022). In 2019, one in eight people worldwide were living
with a mental health condition, with the highest disability-adjusted
life years (DALYs) due to mental and addictive disorders present in
high-income countries such as those in Northern Europe, North America,
and Australia (Rehm \& Shield, 2019; World Health Organization, 2022).
Despite growing awareness, governments have struggled to design
effective responses. Mental health outcomes arise from complex,
multi-level dynamics between the social, economic, and spatial
structures of urban environments---making both diagnosis and
intervention design deeply challenging (Van Der Wal et al., 2021).

Recent work has shifted focus from individual-level vulnerability to
broader social and structural determinants. While high-income nations
show elevated overall DALYs, within-country disparities reveal that
income inequality is a significant predictor of mental health burden
(Rehm \& Shield, 2019). Moreover, precarious working conditions, housing
instability, and neighborhood disadvantage have all been linked to
increased risk of depression and anxiety (Fone et al., 2014; Pevalin et
al., 2017; Rönnblad et al., 2019; Rugulies et al., 2023). This growing
literature highlights that poor mental health is not merely a personal
issue but one rooted in sustained socioeconomic stress and structural
uncertainty.

Building on this, recent studies have introduced the concept of
precariousness as a multidimensional condition characterized by
instability and lack of control across several life domains, including
employment, finances, housing, social relations, and cultural belonging
(Elsenburg et al., 2025; McKee et al., 2017). This broader view reveals
how different forms of insecurity can co-occur and compound, forming an
ecosystem of risk. However, despite growing evidence of association
between precariousness and mental health, a fundamental question remains
unresolved: \emph{How do these factors causally influence one another?}
Most existing studies rely on cross-sectional correlations, leaving open
questions about directionality and feedback. Without stronger causal
insight, it remains difficult to identify priority targets for
intervention or to anticipate and quantify intended and unintended
effects.

This study aims to investigate the causal mechanisms linking
precariousness and depression and how those mechanisms shape system
responses to external intervention. Specifically, we ask: \emph{How does
the internal configuration of the precariousness--depression
relationship affect the outcomes of stress-reducing interventions?} To
answer this, we combine two approaches. First, we use cycle-capable
causal discovery algorithms to infer directional relationships among
domain-specific precariousness factors and depressive symptoms. By
examining both aggregated measures (depression sum score and a composite
precariousness index) and disaggregated variables (individual depressive
symptoms and specific life-domain indicators), we capture broad patterns
of relationship as well as fine-grained causal pathways. Second, we
translate these insights into a computational dynamical model that
simulates how depression and precariousness co-evolve under varying
levels of external stress. This model enables us to probe how system
properties---such as feedback strength and stochastic noise---govern
responsiveness to intervention. When the effects of depression symptoms
onto the social precarity subsystem are weaker than the reverse path,
there is higher sensitivity of depression to intervention, reflecting
both a faster reduction post financial intervention but also quicker
relapse once financial stresses are reestablished. By integrating causal
discovery with dynamic simulation, we demonstrate that intervention
strategies can be tailored to the underlying architecture of
vulnerability.

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study
(Snijder et al., 2017; Stronks et al., 2013). It captures the diverse
population of the city of Amsterdam by including the six main ethnic
groups and provides comprehensive health and lifestyle data, including
depressive symptoms as measured by the PHQ-9 (Galenkamp et al., 2017).
To operationalize indicators of precariousness, we draw on the framework
outlined in previous research (Elsenburg et al., 2025) and select a set
of relevant variables. To ensure a robust representation of
precariousness in our causal discovery models, we conducted exploratory
analyses to identify consistent and meaningful data-driven factor
structures. These analyses led to the identification of five distinct
precariousness factors: three reflecting structural
domains---employment, housing, and social relationships---and two
capturing recent life stressors---financial and relational. Each factor
is composed of multiple observed variables, as outlined below.
Additional details on the HELIUS study and the factor extraction process
are provided in the \hyperref[sec-appendix]{Appendix}.

@@

Following data cleaning and normalization procedures, the analytical
sample comprises 21,628 individuals. In addition to the five identified
precariousness factors, we computed an overall precariousness score by
summing these five factor scores, thereby capturing a composite measure
of accumulated precariousness across domains. Depression is represented
using the PHQ-9 (Kroenke et al., 2001), both as a total sum score and as
individual symptom scores. In the subsequent causal discovery analysis,
we explore the relationship between depression and precariousness using
both aggregated scores and their disaggregated components. See
Figure~\ref{fig-dist} for the overall distributions of the variables
used in the analysis.

@@

To uncover directional relationships between precariousness and
depression, we use constraint-based causal discovery algorithms that
allow for feedback loops and latent confounding. Specifically, we apply
the \emph{Fast Causal Inference} (FCI) and \emph{Cyclic Causal
Inference} (CCI) algorithms, both of which can detect cycles and
non-acyclic structures (Mooij \& Claassen, 2020; Strobl, 2019). For
reference, we also include the \emph{PC} algorithm, one of the most
widely used methods for acyclic causal discovery (Spirtes et al., 2001).
While our full analysis includes results from three causal discovery
algorithms---FCI, CCI, and PC---we focus on FCI in the main text, as it
offers a strong balance between flexibility and theoretical soundness.
FCI accommodates latent confounding and potential feedback, making it
well-suited for the complexity of our data (Mooij \& Claassen, 2020).
The PC algorithm is included for reference but relies on more
restrictive assumptions, such as acyclicity and no unmeasured
confounding. CCI, while capable of modeling cycles explicitly, has its
own theoretical limitations; for example, its output may not fully
preserve the d-separation properties of the true causal graph (Strobl,
2019). Results from both PC and CCI are provided in the
\hyperref[sec-appendix]{Appendix}
(Section~\ref{sec-cci}, Section~\ref{sec-pc}) for completeness and
comparison. For readers interested in detailed explanations of the
algorithms, edge interpretations, and graph types (e.g., PAG, MAAG,
CPDAG), as well as algorithmic assumptions and limitations, please refer
to the \hyperref[sec-appendix]{Appendix}
(Section~\ref{sec-causalprimer}).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Within each condition} (fixed \(\alpha\), threshold, CI test,
  and algorithm), we identify the most frequently occurring edge type
  (e.g., arrowhead, circle, none) across bootstraps for each pair of
  variables (Step 2 in Figure~\ref{fig-workflow}).
\item
  \textbf{Across conditions}, we then determine the most dominant edge
  type based on its frequency. In cases where no edge type is clearly
  dominant (i.e., a tie), we mark the edge as uncertain, represented
  with a dashed line in the final graph (Step 3 in
  Figure~\ref{fig-workflow}).
\end{enumerate}

This summarization process is repeated separately for each causal
discovery algorithm (FCI, CCI, PC), yielding a single robust graph per
algorithm. These final graphs represent only the most stable and
consistent relationships across a wide range of plausible
parameterizations, enhancing both reliability and interpretability. To
support transparency, each final graph is accompanied by a matrix
summarizing the relative frequency of each edge type for every variable
pair, enabling readers to assess the degree of uncertainty or agreement
underlying each connection.

@@

As illustrated in Figure~\ref{fig-concept}, the model includes three
components: social precariousness \(P_i(t)\), depressive symptoms
\(D_i(t)\), and a static individual-specific external stressor \(S_i\),
representing financial stress. Depression and precariousness are modeled
as coupled state variables that influence each other over time, with
\(S_i\) acting as a shared input. These dynamics are formalized as a
pair of linear stochastic differential equations (SDEs):

\[
\begin{aligned}
dP_i(t) &= \lambda_P \big( \alpha_{PS} S_i + \alpha_{PD} D_i(t) - P_i(t) \big) \, dt + \sigma_P \, dW_{P,i}(t) \\
dD_i(t) &= \lambda_D \big( \alpha_{DS} S_i + \alpha_{DP} P_i(t) - D_i(t) \big) \, dt + \sigma_D \, dW_{D,i}(t)
\end{aligned}
\]

Here, \(D_i(t)\) denotes the depressive state and \(P_i(t)\) denotes the
level of social precariousness for individual \(i\) at time \(t\), with
\(S_i\) representing a static, individual-specific external stressor
(financial stress). The parameters \(\lambda_D\), \(\lambda_P\), as well
as the coefficients \(\alpha\) and noise terms \(\sigma\), are assumed
to be constant across individuals and time. The parameters \(\lambda_D\)
and \(\lambda_P\) govern the timescales over which \(D_i(t)\) and
\(P_i(t)\) adjust toward their respective input-driven target states.
The coefficients \(\alpha_{DS}\) and \(\alpha_{PS}\) quantify the direct
effects of the external stressor \(S_i\) on depression and
precariousness, respectively, while \(\alpha_{DP}\) and \(\alpha_{PD}\)
capture the mutual influence between the two dynamic processes.
Stochastic fluctuations are introduced via the independent Wiener
processes \(dW_{D,i}(t)\) and \(dW_{P,i}(t)\), scaled by their
respective volatility parameters \(\sigma_D\) and \(\sigma_P\). This
linear structure offers a transparent and analytically tractable
framework for modeling how exogenous variation in financial stress may
propagate through individual dynamics of depression and precariousness.
Crucially, all parameters can be calibrated directly from empirical
covariances and variances observed in the HELIUS cohort. The resulting
steady-state distribution of the model aligns with empirical measures:
PHQ-9 total score (PHQsum) for depression, and the derived social
precariousness factor (P.soc) for social precarity. The full calibration
procedure is described in Section~\ref{sec-calibration}.

@@

\includegraphics[width=0.4\textwidth,height=\textheight]{img/conceptmodel.pdf}

@@

\caption{\label{fig-concept}High-level schematic of the dynamic model
structure used in simulation. Arrows represent influence coefficients
\(\alpha\), matching the terms in the system of stochastic differential
equations. \(D_i(t)\) and \(P_i(t)\) denote depressive symptoms and
social precariousness for individual \(i\), with \(S_i\) representing
static financial stress.}

@@

We fix the adjustment rates \(\lambda_D = \lambda_P = 1\), which
normalizes the time scale given the cross-sectional nature of the HELIUS
data. To resolve the remaining degree of freedom, we treat the feedback
from precariousness to depression (\(\alpha_{DP}\)) as a free parameter,
and we solve algebraically for all other coefficients --- specifically
the direct effects of stress (\(\alpha_{DS}, \alpha_{PS}\)), the reverse
feedback from depression to precariousness (\(\alpha_{PD}\)), and the
noise terms (\(\sigma_D, \sigma_P\)). These solutions are expressed as
closed-form functions of the empirical covariances: \(\mathrm{Var}(D)\),
\(\mathrm{Var}(P)\), \(\mathrm{Var}(S)\), \(\mathrm{Cov}(D, P)\),
\(\mathrm{Cov}(D, S)\), and \(\mathrm{Cov}(P, S)\).

This structure creates a family of analytically calibrated models, each
consistent with the same observed covariances but differing in their
internal configuration. We constrain the space of valid solutions by
imposing that all structural parameters must be positive, reflecting the
assumption that stress and precariousness contribute adversely to
depression. This constraint restricts the allowable range of
\(\alpha_{DP}\) to approximately \([0, 0.698]\) (see Appendix
Section~\ref{sec-analcalibration}).

Figure~\ref{fig-tradeoff} illustrates how the remaining parameters
adjust as \(\alpha_{DP}\) increases to maintain consistency with the
empirical covariances. As \(\alpha_{DP}\) increases, \(\alpha_{PD}\)
tends to decrease, reflecting a compensatory trade-off between the
mutual influences of depression and precariousness. This interplay helps
preserve the model's match with observed covariances, without requiring
major shifts in external input strength. Meanwhile, the external
coupling terms --- \(\alpha_{DS}\) and \(\alpha_{PS}\) --- also shift to
absorb variance in a way that maintains alignment with the empirical
data. Notably, the noise amplitudes \(\sigma_D\) and \(\sigma_P\)
decline slightly as \(\alpha_{DP}\) increases. This indicates that more
of the observed variability is captured by deterministic interactions,
with less left to be explained by stochastic fluctuations.

@@

In each simulation, financial stress of individual \(i\) (\(S_i\)) is
proportionally reduced toward the lowest observed value, modeling a
population-wide intervention applied at varying intensities (0--100\%).
For each model, we simulate 1000 virtual individuals whose baseline
stress levels are sampled from the empirical distribution and scaled
according to the intervention level.

@@

The causal graph in panel (a) highlights key pathways in the
relationships between precariousness factors and depression
(\emph{PHQsum}). Employment precariousness (\emph{P.emp}) and social
precariousness (\emph{P.soc}) are not identified as causes of
depression, whereas financial stress (\emph{S.fin}) appears to play a
causal role. While \emph{P.emp} and \emph{P.soc} are generally not
recognized as causes of other precariousness factors, \emph{S.fin}
emerges as a potential cause, as indicated by its circle edge endpoint.
This pattern is further supported by the edge-type proportion matrix
shown in panel (b), which aggregates edge inference outcomes across
bootstrapped datasets.

@@

Relational stress (\emph{S.rel}) plays a more nuanced role, interacting
with depression and \emph{S.fin} through a latent confounder. Its
connection to \emph{P.soc} is less definitive---the proportion matrix in
Figure~\ref{fig-sum-2} suggests that this relationship is frequently
absent, ambiguous, or potentially bidirectional across bootstrap
samples. Meanwhile, housing precariousness (\emph{P.hou}) is not
directly causally related to depression but is linked to \emph{P.emp}
and \emph{S.fin}. While \emph{P.emp} and \emph{S.fin} are identified as
non-causes of \emph{P.hou}, it remains unclear whether \emph{P.hou}
causally influences \emph{P.emp} or if their relationship is mediated by
an unobserved confounder.

@@

The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
largely consistent with those from the FCI analysis, with one key
distinction: CCI shows greater ambiguity in the directional relationship
between \emph{S.fin} and \emph{PHQsum}: CCI assigns an arrowhead from
\emph{S.fin} to \emph{PHQsum} in roughly 50\% of bootstraps, while the
remaining cases display an undirected (circle) endpoint. Overall, CCI
introduces greater uncertainty in edge orientations relative to FCI,
yielding more undirected endpoints and more frequent bidirectional
(confounded) edges, which in turn contribute to a slightly higher
overall count of arrowheads when directions are inferred. Despite this,
the skeleton structure and overall causal directions derived from CCI
align well with the results of the FCI algorithm. For example, both
algorithms identify directional pathways from the stressors
(\emph{S.rel}, \emph{S.fin}) to \emph{P.soc}, and suggest that
\emph{P.soc} is not a cause of \emph{PHQsum}, but more likely affected
by it. \emph{P.emp} is not inferred to cause any of its neighboring
variables (\emph{P.hou}, \emph{PHQsum}, \emph{S.fin}), and instead
appears to be affected by them. \emph{P.hou}, in turn, is connected to
\emph{P.emp} and \emph{S.fin} with circle endpoints, indicating that it
may be a potential source of influence rather than an outcome. These
consistent pathways underscore the robustness of the core relational
architecture across algorithms amid variation in directional confidence.

@@

\caption{\label{fig-sum}Causal relationships between five precariousness
factors and depression sum score (\emph{PHQsum}), as estimated using the
FCI algorithm across bootstrapped samples. \textbf{(a)} Aggregated
causal graph showing the most frequent edge types across bootstraps.
Dashed edges indicate ties, with a bold horizontal bar marking where the
tie occurs. \textbf{(b)} Proportion matrix summarizing edge types across
all bootstraps: each matrix cell \([i, j]\) indicates how often a given
edge from variable \(i\) to variable \(j\) exhibited a particular edge
mark (e.g., arrowhead, tail, or circle) adjacent to \(j\). Green
indicates arrowheads, coral arrowtails, blue circle marks, and gray
absence of an edge. Colors are blended and shaded by frequency, with
dominant symbols appearing more opaque. For example,
matrix{[}\(\mathit{S.fin}\), \(\mathit{PHQsum}\){]} shows a green
arrowhead (\(>\)) and matrix{[}\(\mathit{PHQsum}\),
\(\mathit{S.fin}\){]} shows a coral arrowtail (\textemdash), together
indicating that \(\mathit{S.fin} \rightarrow \mathit{PHQsum}\) in the
majority of bootstrapped graphs.}

@@

Moving from the sum score representation to the disaggregated
symptom-level graph provides a more granular perspective on the causal
relationships between precariousness factors and depression symptoms.
Unlike the sum score graph, which aggregates all symptoms into a single
measure, the symptom-level graph reveals the heterogeneity in how
specific symptoms (\emph{con} (concentration), \emph{slp} (sleep),
\emph{ene} (energy), \emph{app} (appetite), \emph{mot} (motor),
\emph{sui} (suicidal), \emph{anh} (anhedonia), \emph{glt} (guilt), and
\emph{dep} (depressed mood)) are influenced by, and in turn influence,
different forms of precariousness.

@@

This discrepancy likely arises from a combination of factors. As the
analysis moves from aggregated to disaggregated variables, the number of
CI tests increases substantially, leading to a loss of statistical
power. With more variables being conditioned on, the effective sample
size per test diminishes, making it harder to detect subtle
dependencies. Over-conditioning---controlling for too many
variables---can also obscure real relationships, resulting in false
negatives. Beyond these statistical power issues, the underlying
assumptions of the two CI tests contribute to their divergence. RCoT is
a nonparametric test that captures both linear and nonlinear
dependencies without assuming a specific distributional form. However,
it is more conservative and requires larger sample sizes, especially in
high-dimensional settings. In contrast, the Gaussian CI test is based on
partial correlations and assumes linear Gaussian relationships, making
it more permissive and more likely to detect weak or spurious
associations. As a result, it is common to see edges where Gaussian CI
suggests a directional relationship, while RCoT indicates no edge. In
the proportion matrix, this appears as a 50--50 tie between absence and
arrowhead symbols.
